Human capital mediates natural selection in contemporary humans

### 2021-07-23 why-natural-selection.Rmd
Describe within-sibling results.

* Add CIs - done.
* Clarify collider bias/endogeneity - done.
* Formalize theory - done.
  - Comment on results where effects are actually reversed. This works
    nicely if you throw in a money cost. Done computational example.
* Clarify causality?
  - Helps to deal with the "collider bias" argument of reviewer
  - Might also be a better test of mediation via human capital
  - One other possibility. If we use only people with first profession
    teacher, that (as it happens) shuts down the EA3-income correlation.
    It also shuts down the EA3-fertility relationship, suggesting that
    income is indeed the mediator.
* Control for PGS for fertility
* Clarify "what we learn about genetic links to fertility by using economic theory to interpret relationships between polygenic scores and fertility."
  - Point is that we can explain how natural selection is occurring.
* "Discuss empirical predictions for AFLB grounded in dynamic theory"
  - hmm, tricky, maybe we can include it in discussion of economic results at
  end of formal section.
* Why don't women see stronger effects? But note, they do: not significant
  after correction for multiple testing, but a 25% difference. 
  - I think we leave this for reviewers.
* Quadratic regressions for parents - DONE

@@

Second, we re-estimate equation \@ref(eq:regression) for each polygenic score,
controlling for earnings in first job and education levels. Effect sizes are
generally reduced (Appendix Figure \@ref(fig:plot-res-earn-educ)), with
large reductions for polygenic scores for educational attainment. This 
suggests that earnings and education indeed mediate the effect of 
polygenic scores on fertility.

@@

cor_pool_within <- cor.test(res_all$estimate.x, res_all$estimate.y)
```

One interpretation of these results is that natural selection on polygenic
scores is driven by their correlation with human capital, which affects
fertility as our model describes. This does not necessarily imply that polygenic
scores are causal of either human capital or fertility. Instead they may simply
correlate with different family environments, for example via "genetic nurture"
effects [@kong2018nature] or population stratification. To test whether
polygenic scores cause fertility, we estimate equation \@ref(eq:regression)
among full siblings, controlling for sibling group fixed effects (Appendix
Figure \@ref(fig:plot-res-fe-fertility)). Results are never significant after
Bonferroni correction. The correlation between pooled and within-sibling effect
sizes is `r cor_pool_within$estimate` (p = `r cor_pool_within$p.value`), and within-sibling effects are about 1/3 smaller. Thus, there is only weak evidence for causal
links from polygenic scores to fertility, with at least some of the effect
coming from gene-environment correlation. If we re-estimate the within-sibling
regressions adding a control for education, effect sizes barely change. This
suggests that our measures of education only mediate between-sibling-group
differences in fertility; among siblings, other causal channels are more
important.

* Results:
  - in bivariate within-sib regs, extraversion, EA2/3, ADHD and some others are
    weakly significant (only at .05!)
  - adding education as a control barely changes effect sizes
  - adding income as a control increases effect sizes (of EA at least)
  - by contrast, in pooled regs, adding education shrinks effect sizes, and
    adding income does nothing.
  - relevant: effect of EA2 on education is halved within-sib, and effect of
    education on fertility is halved within-sib
  - whereas, effect of EA2 on income is halved within-sib, but effect of income
    on fertility is multiplied by 10!! within-sib. Much of this is because
    of year of birth confound i.e. declining fertility and also retired people
    have smaller income
  - within-sib and pooled coefs for PGS are highly correlated; within coefs
    are about 67% of the size.

* Interpretation:
  - some families are better educated and have fewer children
  - they also have selected-against PG scores like what we see
  - scores aren't necessarily causal for children, they may be
    associated
  - the between-family effect goes through education, but not the 
    within-family effect (cf. the paper on causality)

@@

An alternative theory is that polygenic scores correlate with the motivation to
have children (parameter $a$ in the model; cf. @jones2008fertility). However, this
theory would not explain why our results are weaker at higher incomes and
education levels.[^note] Another alternative is that traits under selection are
linked to externalizing behaviour and risk-seeking. This might be partially
captured by our risk-aversion parameter $\sigma$, which affects fertility by
changing the marginal utility of income, but a more direct channel is risky
sexual behaviour [@mills2021identification]. The data here provide some support
for this: scores which might plausibly be linked to externalizing behaviour,
like ADHD and (younger) age at smoking initiation, are selected for. However,
this theory is less good at explaining variation in selection across the full
range of scores, including physical measures such as waist-hip ratio and BMI. We
test this theory directly by re-estimating equation \@ref(eq:regression)
controlling for a measure of risk attitude (UK Biobank field 2040).  The median
ratio of effect sizes between regressions with and without controls is 
`r median(ratio)`; all scores which are significant at $p < 0.05/33$ in
uncontrolled regressions remain so when controlling for risk attitude. This
non-result could simply reflect the imprecision of the risk aversion measure,
which is a single yes/no question. However, the risk aversion measure *does*
predict the overall number of children, highly significantly ($p < 2 \times
10^{-16}$, `r n_risk_sig`/33 regressions). Given that, and the statistical power
we get from our sample size, we believe that the non-result is real: while risk
aversion does predict fertility in the sample, it is not an important channel
for natural selection.

@@

```{r plot-res-earn-educ, fig.cap = "Selection effects controlling for earnings in first job (estimated by mean earnings from ASHE 2007 for the SOC2000 job code) and education (left education before 16, 16-18, or after 18). Raw effects are shown for comparison."}

@@

```{r plot-res-fe-fertility, fig.cap = "Selection effects controlling for sibling-group fixed effects. \"Controlled\" regressions add a control for education (left education before 16, 16-18, or after 18)."}

>>>>

* Only use education, not first job pay, in "controlled" regressions; check
  description is correct - done.
* Add CIs - done.
* Clarify collider bias/endogeneity - done.
* Formalize theory - done.
  - Comment on results where effects are actually reversed. This works
    nicely if you throw in a money cost - done computational example.
* Clarify causality? - done...
  - Helps to deal with the "collider bias" argument of reviewer
  - Might also be a better test of mediation via human capital
  - One other possibility. If we use only people with first profession
    teacher, that (as it happens) shuts down the EA3-income correlation.
    It also shuts down the EA3-fertility relationship, suggesting that
    income is indeed the mediator.
* Clarify "what we learn about genetic links to fertility by using economic theory to interpret relationships between polygenic scores and fertility."
  - Point is that we can explain how natural selection is occurring.
* "Discuss empirical predictions for AFLB grounded in dynamic theory"
  - hmm, tricky, maybe we can include it in discussion of economic results at
  end of formal section.
  - think our theory with 2 periods probably covers this
* Why don't women see stronger effects? But note, they do: not significant
  after correction for multiple testing, but a 25% difference. 
  - I think we leave this for reviewers.
* Quadratic regressions for parents - DONE

@@

Second, we re-estimate equation \@ref(eq:regression) for each polygenic score,
controlling for education levels. Effect sizes are
generally reduced (Appendix Figure \@ref(fig:plot-res-educ)), with
large reductions for polygenic scores for educational attainment. This 
suggests that education indeed mediates the effect of polygenic scores on fertility.

@@

lm_pool_w <- tidy(lm(estimate.y ~ estimate.x, res_all))
coef_pool_w <- lm_pool_w[[2, "estimate"]]
se_pool_w <- lm_pool_w[[2, "std.error"]]
```

One interpretation of these results is that natural selection on polygenic
scores is driven by their correlation with human capital, which affects
fertility as our model describes. This does not necessarily imply that polygenic
scores are causal of either human capital or fertility. Instead they may simply
correlate with different family environments, for example via "genetic nurture"
effects [@kong2018nature] or population stratification. To test whether
polygenic scores cause fertility, we estimate equation \@ref(eq:regression)
among full siblings, controlling for a full set of sibling group dummies
(Appendix Figure \@ref(fig:plot-res-fe-fertility)). All within-sibling effects
are insignificant after Bonferroni correction. Effect sizes are positively
correlated with effect sizes from the pooled model, and about 1/3 smaller
(regressing within-sibling on pooled effect sizes, b = `r coef_pool_w`, se = 
`r se_pool_w` ). Thus, there is only weak evidence that polygenic scores are
causally linked to fertility, and selection effects are at least partially
driven by gene-environment correlation. Also, if we add a control for education 
to the within-sibling regressions, effect sizes barely change. This
suggests that our measure of education only mediates differences in fertility
between sibling groups; among siblings, other causal channels are more
important.

@@

An alternative theory is that polygenic scores correlate with the motivation to
have children (parameter $a$ in the model; cf. @jones2008fertility). However, this
theory would not explain why our results are weaker at higher incomes and
education levels.[^note] Another alternative is that traits under selection are
linked to externalizing behaviour and risk-seeking. This might be partially
captured by our risk-aversion parameter $\sigma$, which affects fertility by
changing the marginal utility of income, but a more direct channel is risky
sexual behaviour [@mills2021identification]. The data here provide some support
for this: scores which might plausibly be linked to externalizing behaviour,
like ADHD and (younger) age at smoking initiation, are selected for. However,
this theory is less good at explaining variation in selection across the full
range of scores, including physical measures such as waist-hip ratio and BMI. We
test this theory directly by re-estimating equation \@ref(eq:regression)
controlling for a measure of risk attitude (UK Biobank field 2040).  The median
ratio of effect sizes between regressions with and without controls is 
`r median(ratio)`; all scores which are significant at $p < 0.05/33$ in
uncontrolled regressions remain so when controlling for risk attitude. This
non-result could simply reflect the imprecision of the risk aversion measure,
which is a single yes/no question. However, the risk aversion measure *does*
predict the overall number of children, highly significantly ($p < 2 \times
10^{-16}$ in `r n_risk_sig`/33 regressions). Given that, and the statistical power
we get from our sample size, we believe that the non-result is real: while risk
aversion does predict fertility in the sample, it is not an important channel
for natural selection.

@@

```{r plot-res-educ, fig.cap = "Selection effects controlling for education (left education before 16, 16-18, or after 18). Raw effects are shown for comparison."}

@@

```{r plot-res-fe-fertility, fig.cap = "Selection effects controlling for sibling-group fixed effects, with and without a control for education (left education before 16, 16-18, or after 18)."}
